\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{paper-2026-acsos-phd-symposium}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
%\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Towards Collective Robotic Operating Systems through Aggregate Computing\\
}

\author{Angela Cortecchia\\
\IEEEauthorblockA{\textit{Department of Computer Science and Engineering} \\
\textit{University of Bologna}\\
Cesena, Italy \\
angela.cortecchia@unibo.it}
}

\maketitle

\begin{abstract}
Swarm robotic systems are increasingly relevant for missions in which many heterogeneous devices must operate in dynamic and potentially safety-critical environments.
%
Programming such systems remains challenging:
developers need to express objectives at the collective level,
while their execution is distributed across devices that may move,
fail, join or leave the system,
without a centralized point of coordination.
%
\acl{ac} addresses this issue by providing macroprogramming abstractions for specifying resilient and self-organizing collective behaviors.
%
However,
current \acl{ac} applications typically execute a single collective program and offer limited support for runtime adaptation,
preemption, concurrent applications, and safety guarantees during transient phases.

This research investigates the foundations of a \acl{cros} built on \acl{ac}:
a middleware layer that lifts ROS-inspired capabilities from individual robots to the collective level.
%
The paper outlines the main requirements for such a system and reports ongoing contributions on morphogenetic formation control,
runtime path replanning, self-stabilizing consensus, collective sensing, and safety guarantees.
%
These results provide initial building blocks towards a collective operating layer for swarm robotics,
with future work focusing on safe runtime adaptation and support for multiple concurrent collective processes.
\end{abstract}

\begin{IEEEkeywords}
aggregate computing, distributed systems, self-organizing systems, collective operating systems
\end{IEEEkeywords}

\section{Overall Problem and Motivation}\label{sec:overall-problem-and-motivation}
Swarm robotics are characterized by the presence of a large number of heterogeneous devices,
which are mobile and resource-constrained.
%
An example is a network of surveillance drones,
which have to coordinate among each other to monitor a certain area,
while maintaining a certain swarm formation, avoiding collisions and respecting no-fly zones
and possibly allowing law enforcement to intervene in case of emergency.
%
Also environmental monitoring systems and search-and-rescue operations can benefit from swarm robotics,
such as wildfire monitoring and disaster response.

Those devices must coordinate among each other to achieve global objectives,
while adapting to changes in the environment and in the network topology,
possibly without a centralized point of coordination,
which can introduce a single point of failure and limit the scalability of the system.
%
Those systems are often deployed in dynamic and unpredictable environments,
where devices may fail or be added/removed at any time.

The design of such systems is challenging,
as it requires to ensure that the system can autonomously adapt to changes in the environment and in the network topology,
while still achieving its goals.

A possible approach to design swarm robotics systems is through macroprogramming~\cite{casadei22},
which allows developers to focus on the desired global behavior of the system,
rather than on the individual behavior of each device.
%
On top of this concept,
is built the \ac{ac} paradigm~\cite{DBLP:journals/computer/BealPV15},
which provides a programming model that allows to express the collective-level behavior of the system through composition of \emph{computational fields}~\cite{JLAMP2019},
a distributed data structure that maps devices to their values,
and that can be manipulated through a set of high-level operators.

By design,
\ac{ac} is resilient to failures and scalable to large numbers of devices,
as it relies on local interactions and \emph{self-organization} to achieve global objectives.
%
Over the years, reusable building blocks~\cite{DBLP:journals/tomacs/ViroliABDP18}
supporting self-stabilizing~\cite{DBLP:journals/cacm/Dijkstra74} \ac{ac} behaviors
have been proposed and later used in high-level libraries such as MacroSwarm~\cite{DBLP:journals/lmcs/AguzziCV25},
which supports pattern formation in swarm robotics.
%%
%Moreover,
%a general formal framework for the specification of collective computational processes has been proposed~\cite{EAAI2020-processes},
%which are temporary distributed computations carried out by dynamic groups of devices,
%as well as a set of algorithms for efficient propagation and termination of such concurrent execution processes~\cite{DBLP:journals/fgcs/AudritoCT24}.

However,
typical \ac{ac} applications run a single although complex algorithm on each device,
without preemption or possibility of changing dynamically the behavior of the system at runtime,
and no users permission definition is provided.
%
In the context of swarm robotics,
this is a limitation,
as multiple applications may need to run concurrently on the same devices,
and potentially third-party users (e.g., law enforcement) may need to stop or change the behavior of the system at runtime in case of emergency.
%
Moreover, %% todo ale
\ac{ac} leverages on the \emph{eventual} consistency of the system to achieve self-stabilization,
which may not be sufficient in safety-critical scenarios,
where a temporary inconsistency may lead to unsafe situations,
such as surveillance drones colliding with each other or entering restricted areas.

\section{Objectives and Contributions}\label{sec:objectives-and-contributions}
% todo Obiettivo -> presentare un sistema operativo collettivo per gli sciami basato su macro-programmazione
% todo Perche? -> paradigma con astrazioni globali, dec, ..

This work aims to address \ac{ac} gaps in the context of swarm robotics,
by proposing a set of mechanisms and algorithms that can be used to build the base of a ROS-like~\cite{ros} \ac{cros} based on \ac{ac},
and by providing a set of reusable self-organizing building blocks that support easy global objectives definition and self-organization.
%
The challenges in designing and implementing such systems are manifold,
as they require to ensure that the system can autonomously adapt to changes in the environment and in the network topology,
while still achieving its goals.

The requirements for such swarm systems include
\begin{inlinelist} %%add motivation
    \item the support for concurrent execution of multiple applications on the same device, parallelizing the execution of different tasks;
    \item preemption and runtime mechanism for dynamic changes in the behavior of the system, to allow third-party users to change the behavior of the system at runtime in case of emergency;
    \item mechanisms to ensure safety in the stabilization phase of the system, to avoid unsafe situations;
    \item heterogeneous devices that act as a collective sensor, to gather distributed information from the environment;
    \item collective-aware team action and formations, to coordinate the movement of the devices in a swarm and to maintain a certain formation;% todo aggiugi formazione spaziali per il movimento coordinato
    \item consensus and communication mechanisms between process in different spatial regions. %todo meccanismi di consenso e communicazione tra processi in regioni spaziali differenti, ad esempio uno spazio di memoria condiviso.
\end{inlinelist}

\section{Methods and Results}\label{sec:methods-and-results}

The research work to cover those gaps is currently ongoing,
and it is being carried out through a bottom-up approach,
starting from specific scenarios and use cases that aim to address individual gaps,
and then generalizing the solutions to be applicable to a wider range of scenarios.

% per prima cosa dovevamo scegliere un linguaggio,
% negli anni ne hanno fatti tot, abbiamo scelto ck perche portabile, multiplatform, offre una libreria ergonomica

As a first step,
the choice of the programming language to use for the implementation of the proposed solutions has been made.
%
During the years,
different \acp{dsl} have been proposed to support \ac{ac} programming,
such as Protelis~\cite{DBLP:journals/tocl/AudritoVDPB19},
Scala-based ScaFi~\cite{DBLP:journals/softx/CasadeiVAP22},
Python-based Phyelds~\cite{DBLP:journals/corr/abs-2603-29999},
and the Kotlin Multi-Platform based Collektive~\cite{DBLP:conf/acsos/Cortecchia24}.
%
In this work,
we use \emph{Collektive} as the reference \ac{ac} language,
for its portability across platforms and its ergonomic library support for implementing \ac{ac} programs.

To support collective formation and spatial organization,
we proposed a \emph{morphogenetic} approach to structure formation~\cite{DBLP:journals/comintsys/CortecchiaCCP26,DBLP:conf/acsos/CortecchiaPCC24}.
%
The approach leverages resource-flow management in a network of devices,
distributing resources towards the most successful area of the network in a self-organizing and self-adaptive way.
%
It is currently being generalized to generate resilient structures that follow mathematically defined shapes.

Runtime adaptability has been investigated through a path-replanning mechanism~\cite{DBLP:conf/acsos/AguzziBCMPPV25}.
%
In this setting,
devices are assigned tasks that must be collectively executed while preserving the global objective.
%
When disruptions occur,
such as the failure of a device,
the swarm adapts its execution plan by redistributing the affected tasks among the remaining devices.

For distributed consensus,
we developed a self-stabilizing \emph{min-max gossip} algorithm~\cite{DBLP:conf/coordination/CortecchiaPV26}.
%
The algorithm provides an adaptive and resilient mechanism for reaching consensus on a selector,
such as the minimum or maximum value of sensed information distributed across the swarm.
%
Unlike classical approaches,
it tolerates failures and topology changes without requiring explicit parametrization to reset the gossip memory.

Collective sensing and monitoring have been addressed through \ac{dpf}~\cite{DBLP:conf/acsos/CortecchiaDCCV26},
an approach for state estimation in distributed systems.
%
By exploiting \ac{ac},
the proposed algorithms improve resilience to failures and support more accurate neighborhood estimation,
without requiring excessive information exchange among devices.

Safety during transient phases is the focus of ongoing work on runtime filtering.
%
This direction leverages \ac{cbf}~\cite{DBLP:conf/eucc/AmesCENST19} and \ac{clf}~\cite{DBLP:journals/ieeejas/LiWYWH23} theory applied to \ac{ac}\footnote{\url{https://github.com/angelacorte/carol-experiments}}. %%todo, add reference to the paper when (if) accepted,
%
\ac{cbf} and \ac{clf} are used to define system-level constraints,
such as minimum or maximum distances between devices and unsafe areas to be avoided.
%
These constraints are then enforced at runtime while adaptive swarm-level objectives are computed.

% Partirei dai requisiti e parlerei dei contributi
% formazione
%The first contribution towards \ac{cros} is a morphogenetic approach to resource flow management in a network of devices,
%called \emph{FieldVMC}~\cite{DBLP:journals/comintsys/CortecchiaCCP26,DBLP:conf/acsos/CortecchiaPCC24},
%which allows to distribute the available resources in the network from a source to the most successful area of the network.

%This approach can be used to build self-organizing and self-adaptive artificial structures,
%resilient to failures and dynamic changes in the environment.

% todo runtime changes
%The second contribution is a runtime path replanning mechanism.

%A set of devices in an environment are assigned with a set of tasks to be executed,
%they have to coordinate among each other to achieve the global objective of executing all the tasks,
%adapting their execution plan to changes such as a device failure,
%meaning that the tasks assigned to the failed device have to be reassigned to other devices in the network.

%This has been achieved by leveraging \emph{path replanning}~\cite{DBLP:conf/acsos/AguzziBCMPPV25},
%and \ac{ac} primitives for coordination and communication.

% todo consensus
%For what it concerns consensus among distributed devices,
%a self-stabilizing \emph{min-max gossip} algorithm has been proposed~\cite{DBLP:conf/coordination/CortecchiaPV26},
%to provide an adaptive and resilient mechanism to reach consensus on a certain selector among a set of distributed devices.

%This algorithm can be used to build self-organizing and self-adaptive systems,
%where devices can reach consensus on a certain value or decision,
%in respect of classic approaches,
%it is resilient to failures and dynamic changes in the environment,
%and it does not need parametrization to reset the gossip memory when the network topology changes.

% todo collective sensing
%From the perspective of collective sensing and monitoring,
%a set of self-organizing algorithms have been proposed through \ac{dpf}~\cite{DBLP:conf/dcoss/CortecchiaDCCPV26,DBLP:conf/acsos/CortecchiaDCCV26},
%an approach for state estimation in distributed systems.

%\ac{ac} has been used to improve the resilience of the system to failures,
%and to provide a better neighboring estimation without the need of exchanging excessive information among devices.

%A safety filter mechanism is ongoing work,
%by leveraging \ac{cbf}~\cite{DBLP:conf/eucc/AmesCENST19} and \ac{clf}~\cite{DBLP:journals/ieeejas/LiWYWH23} theory applied to \ac{ac} \footnote{\url{https://github.com/angelacorte/carol-experiments}} %%todo, add reference to the paper when (if) accepted,
%in order to provide a runtime mechanism to ensure safety in the transient phase of the system and guarantee that the system will converge to a safe state.

%\ac{cbf} and \ac{clf} are used to define a set of constraints on the system,
%which are then used to enforce safety on the system at runtime,
%while computing adaptive swarm-level objectives.

%+ carol \footnote{\url{https://github.com/angelacorte/carol-experiments}} %%todo, add reference to the paper when (if) accepted


\section{Future Work and Research Plan}\label{sec:future-work-and-research-plan}

As next steps of the research,
the focus will be on the safety filter mechanism,
to improve it by defining constraints on swarm formation and runtime target changing policy.
%
A follow-up of this work can be the integration of the \ac{cbf} within the morphogenesis algorithm,
to define the formation of more complex structures.

Moreover,
a particular focus will be on the support of preemption and the execution of multiple applications on the same device,
since this is a fundamental requirement for a \ac{cros} to be used in real-world scenarios.
%
Currently,
a general formal framework for the specification of collective computational processes has been proposed~\cite{EAAI2020-processes},
which are temporary distributed computations carried out by dynamic groups of devices,
as well as a set of algorithms for efficient propagation and termination of such concurrent execution processes~\cite{DBLP:journals/fgcs/AudritoCT24}.
%
However,
this approach has still some limitations on how a computational process should spread or reduce in space,
so this will be investigated in the future to provide a more general approach to the execution of multiple applications on the same device.

Current contributions are not yet integrated into a unified framework;
therefore,
the final goal of this research is to define a prototype of a \ac{cros} based on \ac{ac} and Collektive,
focused on providing a set of reusable building blocks and algorithms that support the definition of global objectives and self-organization.

\section*{Acknowledgment}
This work has been supervised by Prof. Danilo Pianini and co-supervised by Prof. Mirko Viroli.

\bibliographystyle{IEEEtran}
\bibliography{bibliography}

\vspace{12pt}


\end{document}
